---
title: "Dataset Description"
author: "Ignacio Toledo"
date: "03/24/2015"
output: html_document
---

```{r message=FALSE, echo=F}
library(foreign)
library(dplyr)
library(tidyr)
library(knitr)
```

The following code chunk is not executed, but it gives information about loading
the data sets from the SPSS format, converting the string variables to UTF-8 
encoding, and storing the files in RData.

```{r eval=FALSE}
households <- read.spss('~/Downloads/personas_spss.sav', to.data.frame = T)
expenses <- read.spss('~/Downloads/gastos_spss.sav', to.data.frame = T)
cate <- read.spss('~/Downloads/CCIF_VIIEPF.sav', to.data.frame = T)

nhou <- c("PARENTESCO", "TVP", "VS", "SP01", "SP02", "SP03", "AECISE", "CISE_4", 
          "AECIUO", "PETG",  "CAEG", "CISEG", "AEPRL", "EDUNIVEL", "SP04",
          "AECIUO2", "AE01", "AE02", "AE03", "EDUACTUAL", "AE08")
for (i in nhou) {
  tempo <- levels(households[, i])
  tempo <- iconv(tempo, 'latin1', 'utf-8')
  levels(households[, i]) <- tempo
}

tempo <- levels(expenses$GLOSA)
tempo <- iconv(tempo, 'latin1', 'utf-8')
levels(expenses$GLOSA) <- tempo

save(file='households.RData', households)
save(file='expenses.RData', expenses)
save(file='descript.RData', description_households)
```

## Description of the data sets

```{r}
load('households.RData')
load('expenses.RData')
load('education.RData')
str(households)
str(expenses)
```

The `households` data set contains `r ncol(households)` variables and 
`r nrow(households)` observations, corresponding to a total of 
`r length(unique(households$FOLIO))` households. The data set have
`r t <- 
  as.data.frame(sapply(households, class), columns=c('type')); 
  names(t) <- c('type'); length(t$type[t$type == 'factor'])` factor and 
`r t <- 
  as.data.frame(sapply(households, class), columns=c('type')); 
  names(t) <- c('type'); length(t$type[t$type == 'numeric'])` numeric varibles.



```{r echo=F, eval = FALSE}
hlabels <- as.data.frame(attr(households, 'variable.labels'))
names(hlabels) <- c('Variable Description')
hlabels$"Variable Description" <- c(
  "Zone","Household unique identifier.", "Person order within a household.",
  "Expansion factor.","Household's number of members.",
  "Is the person the Household's head?",
  "Relationship to the households's head.","Person's gender.","Person's age.",
  "Person's marital status of the person.","Person's years of schooling.",
  "Person's number of approved courses.","Person's educational level.",
  "Did the individual finish the educational level?",
  "The person who contributed most to the household budget: Main breadwinner.",
  "Person in charge of household's purchases.","Household type of tenure.",
  "Second houshold type of tenure (if any).","Person's health insurance",
  "Person's retirement found.",
  "International Classification of Status in Employment (ind.)",
  "International Classification of Status in Employment (summary, ind.)",
  "Uniform International Classification of Occupations (ind.)",
  "Person's working population catergory.",
  "Person's economical activity status.",
  "International Classification of Status in Employment, summary (2 cat.)",
  "Reference period of the main economic activity.",
  "Total expenses per household adjusted for non-response (w/o imputed rent)",
  "Total expenses per household adjusted for non-response (with imputed rent)",
  "Total available household income (without imputed rent)",
  "Total available household income (with imputed rent)",
  "Total gross household income (without imputed rent)",
  "Total gross income per household (with imputed rent)",
  "Imputed rent, main household","Imputed rent, second household",
  "Income from dependent employment","Income from self-employment",
  "Available income from salary","Available income from salary imputed by HD",
  "Gross income from salary","Gross income from salary imputed by HD",
  "Available income from ","Disposable income of fees imputed by HD",
  "Gross income from fees observed","Gross income from fees imputed by HD",
  "Income available for those with business on their own observed",
  "Income available for those with business imputed with HD",
  "Gross income of those who are self-employed business noticed",
  "Gross income for those with business imputed with HD",
  "Disposable income of independent professionals observed",
  "Disposable income of the accused freelancers for HD",
  "Gross income of freelancers observed",
  "Gross income of independent professionals imputed with HD",
  "Gross income in retirement observed",
  "Gross income in retirement accused by HD",
  "Gross income from other salaried jobs",
  "Gross income from other freelance jobs","Income Property",
  "Income from financial instruments","Income from net transfers",
  "Identification of home with underreported income")
kable(hlabels,format='markdown', align=c('r'))
```

```{r eval = FALSE}
nh <- names(households)
nh[42:76]
for (i in nh[42:76]) {
  households[, i] <- households[, i] / 612
}
```


01-12 - Individual consumption expenditure of households
01 - Food and non-alcoholic beverages
01.1 - Food
01.2 - Non-alcoholic beverages
02 - Alcoholic beverages, tobacco and narcotics
02.1 - Alcoholic beverages
02.2 - Tobacco
02.3 - Narcotics
03 - Clothing and footwear
03.1 - Clothing
03.2 - Footwear
04 - Housing, water, electricity, gas and other fuels
04.1 - Actual rentals for housing
04.2 - Imputed rentals for housing
04.3 - Maintenance and repair of the dwelling
04.4 - Water supply and miscellaneous services relating to the dwelling
04.5 - Electricity, gas and other fuels
05 - Furnishings, household equipment and routine household maintenance
05.1 - Furniture and furnishings, carpets and other floor coverings
05.2 - Household textiles
05.3 - Household appliances
05.4 - Glassware, tableware and household utensils
05.5 - Tools and equipment for house and garden
05.6 - Goods and services for routine household maintenance
06 - Health
06.1 - Medical products, appliances and equipment
06.2 - Outpatient services
06.3 - Hospital services
07 - Transport
07.1 - Purchase of vehicles
07.2 - Operation of personal transport equipment
07.3 - Transport services
08 - Communication
08.1 - Postal services
08.2 - Telephone and telefax equipment
08.3 - Telephone and telefax services
09 - Recreation and culture
09.1 - Audio-visual, photographic and information processing equipment
09.2 - Other major durables for recreation and culture
09.3 - Other recreational items and equipment, gardens and pets
09.4 - Recreational and cultural services
09.5 - Newspapers, books and stationery
09.6 - Package holidays
10 - Education
10.1 - Pre-primary and primary education
10.2 - Secondary education
10.3 - Post-secondary non-tertiary education
10.4 - Tertiary education
10.5 - Education not definable by level
11 - Restaurants and hotels
11.1 - Catering services
11.2 - Accommodation services
12 - Miscellaneous goods and services
12.1 - Personal care
12.2 - Prostitution
12.3 - Personal effects n.e.c.
12.4 - Social protection
12.5 - Insurance
12.6 - Financial services n.e.c.
12.7 - Other services n.e.c.