---
title: "Project 3: EDA"
author: "Ignacio Toledo"
date: "03/15/2015"
output:
  html_document:
    number_sections: yes
    theme: readable
    toc: yes
---

# Introduction

For this project the data from the **"VII Encuesta de Presupuestos Familiares"**
(VII Household Budget Survey) was selected. This is a survey done every 5 years 
in Chile by means of a survey to a representative sample of homes, which allows 
to know the expenditure structure and consumption patterns of the Chilean 
families and individuals. The main use for this survey the construction of a 
list of goods and services whose costs are monitored monthly to calculate the 
"IPC", or consumer price index, which is the main index used to measure 
inflation in Chile. However, this survey is also an important tool that gathers
and collects an import ammount of socioeconomic information for urban households
and their inhabitants, including age, gender, housing tenure, incomes, education
and working conditions.

The data used for this work is splited in two data sets:

* `households` contains 43 variables where each observation (row) represent an
inhabitant. Some of the variables are related to the individuals, like age,
marital status, education, gender, incomes, etc.; other variables are an 
aggregation over the households the inhabitans belong to (`home.id` variable),
summarizing the household's total income and expenses.

* `expenses` has 10 variables, with one observation by each expenditure 
monitored by the survey. All of the variables are categorical, except for the
numericla variable `expense` that records the amount expended in each 
transaction. It also includes the `home.id` variable that allows to join this 
data with the households in the first data set. 

Some data wrangling will be needed before starting some of the exploratory data
analysis,  given that the `households` data set have one entry by each household
member,  while the `expenses` data set contains the expenses only by household 
not separated by household inhabitant: this means that is possible to  merge the
data by matching the household ids, but not by individuals (because that was
the intended use for the data).

# Loading libraries and data set {#load}

The following libraries were used for this work:

* `ggplot2`
* `gridExtra`
* `GGally`
* `ggthemes`
* `dplyr`
* `tidyr`
* `knitr`

```{r libraries, message=F, echo=F}
library(ggplot2)
library(ggthemes)
library(GGally)
library(dplyr)
library(gridExtra)
library(tidyr)
library(knitr)
library(pander)
library(xtable)
library(reshape2)
library(scales)
theme_set(theme_bw())
```

The data is stored in RData files, after being transformed from SPSS data sets.

```{r load_data}
load("households.RData")
load("expenses.RData")
```

Some cleaning is still needed, for example there two negative ages and some
households without a total income reported, because of missing data.

```{r clean_data}
households <- subset(households, age >= 0 & !is.na(income.hh.av.rent))
```

# Univariate Plots and Analysis.

Let's start with some simple explorations of the population in our data set. 
From the variables descriptions we decided to focus on the following variables:

* By individuals:
    + Age (`households$age`).
    + Education Attainment (`households$edu.level`).
    + Relationship with Household Head (`households$kinship`).
    + Type of Educational Establishment, for inhabitants doing studies 
      (`households$edependence`).
    + Health Expenditure (`households$head.exp`).
* By household:
    + Principal Household type of Tenure (`households$tph`).
    + Household's Available Monthly Income, with imputed rent 
      (`households$income.hh.av.rent`).
    + Number of Inhabitants by Household (`households$num.inhabitants`).


## Household's inhabitants Age Distribution

What is the population's age distribution? The `summary` function can give us
a start:

```{r age_sum, echo=F, results='asis'}
t1 <- summary(households[,'age'])
pandoc.table(t(t1), 
             caption = "Age Distribution Summary")
```

So, the average population's age is 32 years old. However, a plot will give us 
much more information about the age distribution of the population surveyed in 
our data set. The next figure shows an histogram using the `age` variable (a 
discrete numerical variable). The binwidths are equal to 1 year. The plot shows 
that the population is not normally distributed and positevely skewed overall. 
This is expected, since the population must decrese with age as people dies by
accidents, illness or natural causes. 

However is interesting to notice some peaks at around 5, 25 and 50 years of age:
they might correspond to generations with higher natality rates or less infant
mortality.


```{r age_hist, fig.align='center', echo=F}
ggplot(aes(age), data = households) + 
  geom_histogram(binwidth = 1, fill = "#67a9cf", color = "grey") +
  xlab("Age") +
  ylab("Freq.") +
  ggtitle("Age Distribution")
```

---

## Educational attainment

The educational attainment measures the educational level attained by an 
individual. In our data set the variable that measures this is called 
`edu.level`, and is a categorical variable stored as a factor with 16 levels
plus a "NA" for missing information. The next figure shows a bar plot with
the distribution of this variable for our population. A logarithmic scale
was used in the Y axis to display better the differences, as some categories
have a small number of cases.

```{r eduatt, fig.align='center', echo=F}

ggplot(aes(edu.level), data = households) + 
  geom_histogram(fill = "#67a9cf") + 
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  scale_y_continuous(trans = 'log10', breaks=c(10, 100, 1000, 5000)) +
  xlab("Educational Attainment") +
  ylab("Freq.") +
  ggtitle("Educational Attainment Distribution")
```

---

## Kinship

The variable `kinship` is a categorical variable stored as a factor vector with
14 levels indicating the kinship relation of the different individuals with 
respect to the household head. We present a figure with a bar plot to analyse
this variable distribution. Notice that again the Y axis scale is logarithmic so
we don't loose information given the small number of cases of some 
relationships.

Most of the inhabitants of a household have a direct relation with the 
household head: they are either the children or the spouse in most cases.

```{r kinship, fig.align='center', echo=F}
# For displaying purposes a character vector is created, that orders the levels
# by Frequency.
ord_kin <- as.character(
  arrange(data.frame(table(households$kinship)), desc(Freq))$Var1)

ggplot(aes(factor(kinship, levels = ord_kin),
           y = ..count..),
       data = households) + 
  geom_histogram(fill = "#67a9cf") + 
  scale_y_continuous(trans = 'log10', breaks = c(10, 100, 1000, 5000)) + 
  theme(axis.text.x=element_text(angle = 40, hjust = 1)) +
  xlab("Relationship with Household Head") +
  ylab("Freq.")
```

---

## Household Tenure

The household tenure studies the kind of tenure a houshold has over its main
dwelling place. `tph` is a categorical variable which again is stored as a 
factor vector, with 9 levels plus a "NA" category for missing information.

The following figure shows a bar plot with the percentage each category 
represents over the total number of households surveyed. Over 60% of the 
households are either fully owned or owned through a mortage still being payed.
I was expecting a higher percentage of households paying a rent, however only
around the 15% of the households fall in this category.

```{r houseten, fig.align='center', echo=F}
# For displaying purposes a charcter vector is created, that orders the levels
# by Frequency
ord_tph <- as.character(
  arrange(data.frame(table(households$tph)),desc(Freq))$Var1)

ggplot(aes(factor(tph, levels = ord_tph),
           y = 100 * ..count.. / sum(..count..)),
       data = households) + 
  geom_histogram(fill = "#67a9cf") +  
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  xlab("Kind of Household Tenure") +
  ylab("Percentage")
```

---

## Household's Income

We analyse here the monthly household's income. A household's income adds up all 
the incomes from the inhabitans of a household, including the incomes from 
dependent work activities, independant works, rents, social helps, financial 
instruments, pentions, etc. Our data set includes 10517 households. As shown in 
the following table, the summary of this variable tell us that the median income 
is US$1110 while the mean is US$1707, and 75% of the households earn less than
US$1976.

```{r houseinc_summ, echo=F, results='asis'}
# We create `houseinc` data frame that contains only one income by household, 
# and reatin only the `home.id` and `income.hh.av.rent` variables
houseinc <- 
  households %>% 
  subset(person.id == 1) %>% 
  select(home.id, income.hh.av.rent)

t1 <- summary(houseinc$income.hh.av.rent)
pandoc.table(t(t1),
             caption = "Avilabe Income by Household Summary")
```

The next figure plots an histogram of the household's income.

```{r houseinc, echo=F, fig.align='center'}
ggplot(aes(income.hh.av.rent), data = houseinc) +
  geom_histogram(fill = "#67a9cf", binwidth = 100) +
  xlab("Household Monthly Income (US$)") +
  ylab("Frequency")
```

---

We can see that most of the data in under US$2000. For the next figure we change
the X axis scale from linear to logarithmic, and we use `geom_density` instead
of `geom_histogram`.

```{r houseinc_log_den, echo=F, fig.align='center'}

ggplot(aes(income.hh.av.rent), data = houseinc) +
  geom_density(binwidth = 0.1, fill = "#67a9cf") +
  scale_x_continuous(
    trans = "log1p",
    breaks=c(3, 420, 576, 739, 908, 1110, 1370, 1710, 2310, 3610, 53300)
    ) +
  xlab("Household Monthly Income (US$)") +
  ylab("Density") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

The plot shows a better description of the household's income. We can see the
peak at around US$1100, which is the median of the income distribution. However
this plot can be misleading for some readers, because it could lead them to
believe that the wealth is well distributed within the chilean's households.
---

So, we plot once again a histogram with linear scales in both axis, but we will
remove the 10% of the households with higher incomes. The next figure shows
the result.

```{r houseinc_det, echo=F, fig.align='center'}

ggplot(aes(income.hh.av.rent), data = houseinc) +
  geom_histogram(binwidth = 100, fill = "#67a9cf") +
  scale_x_continuous(
    breaks = c(3, 420, 576, 739, 908, 1110, 1370, 1710, 2310, 3610),
    limits = c(0, 3610)
    ) +
  xlab("Household Monthly Income (US$)") +
  ylab("Frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

---

The reader can wonder about the 10 kind of arbitrary ticks used in the X axis.
These values represent the limits of the income deciles for the chilean 
household's incomes. The following table ilustrate these deciles.

```{r income_decil, echo=F, results='asis'}

decil <- quantile(
  houseinc$income.hh.av.rent,
  probs = seq(0, 1, 0.1),
  na.rm = T
  )

pandoc.table(
  t(data.frame(Percentil = names(decil),
               'US Dolars' = format(decil, digits = 2))),
  caption = "Table with percentiles calculated to create Income Deciles",
  split.tables = 120,
  digits = 4)

decil[1] <- 0 
decil[11] <- decil[11] + 100
```

These deciles were calculated using the `quantile` function, where we looked
for the limits that would separate the households in 10 different deciles, each
one compromising a 10% of the households. Using this new variable that we 
called `income.dec`, we created a bar plot that is shown in the next figure.
The plot shows how the wealth is distributed in the chilean households: the
decile with the top 10% of households with more income is getting around the 
35% of the total wealth.


```{r income_decil_2, echo=F}
households$income.dec <- 
  cut(households$income.hh.av.rent, decil, right = F)

houseinc$income.dec <- 
  cut(houseinc$income.hh.av.rent, decil, right = F)

levels(houseinc$income.dec) <- c(
  "Up to US$419", "US$420 to US$575", "US$576 to US$738", "US$739 to US$907",
  "US$908 to US$1109", "US$1110 to US$1369", "US$1370 to US$1709", 
  "US$1710 to US$2309", "US$2310 to US$3609", "US$3610 and Higher"
  )

levels(households$income.dec) <- levels(houseinc$income.dec)
```


```{r totincquant_plot, echo=F, fig.align='center'}
indec <- 
  subset(houseinc, !is.na(income.dec)) %>% 
  group_by(income.dec) %>% 
  summarise(Total = sum(income.hh.av.rent, na.rm = T))

ggplot(aes(income.dec, y = 100 * Total / sum(Total)), data = indec) +
  geom_bar(stat = "identity", fill = "#67a9cf") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('Household Incomes\' Deciles') +
  ylab('Percentage of the Total Household Income') +
  ggtitle('Household Total Income Share by Deciles')
```


---

## Inhabitants by household

The variable `num.inhabitants` holds the number of inhabitans by household.
The following figure shows a histogram with the distribution of the inhabitant
number by household.

```{r inhab_sum, echo=F, fig.align='center'}
inhab <- 
  households %>%
  subset(person.id == 1, select = c("home.id", "num.inhabitants"))

ggplot(aes(as.factor(num.inhabitants)), data = inhab) +
  geom_histogram(fill = "#67a9cf") +
  scale_x_discrete() +
  xlab('Number of Inhabitants by Household') +
  ylab('Frequency')
```

From the following summary table we can conclude that the median population
by household is 3, while the mean population is 3.389 inhabitants.

```{r inhab_summ, echo=F, results='asis'}
t1 <- summary(inhab$num.inhabitants)
pander(t(t1))
```

---

## Educational Institutions

In this section we study the variable `edependence`. This variable holds the
kind of educational institution the students in our population are attending.
This variable is categorical, and is stored as factor with 12 levels, including
1 level for the individuals not studying. In the next figure we show the 
distribution ordered by the number of cases by institution, removing the 
population that is not currently attending any institution or studying.

```{r edudep, fig.align='center', echo=F}
ord <- as.character(
  arrange(data.frame(
    table(subset(households, edependence != "Not Studying")$edependence)), 
    desc(Freq))$Var1)

ggplot(aes(factor(edependence, levels = ord)),
       data = subset(households, edependence != "Not Studying")) +
  geom_histogram(fill = "#67a9cf") + 
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  xlab('Educational Institution Type (private/public)') +
  ylab('Frequency')
```

---

Regarding the primary and secundary education, the next figure shows a bar plot
with the percentage of students that are attending either a public school, a
private school or an state subsidized institution.

```{r edudep2, fig.align='center', echo=F}
ggplot(aes(factor(edependence, levels = ord),
           y = 100 * ..count.. / sum(..count..)),
       data = households[households$edependence %in% ord[c(3,2,5)], ]) +
  geom_histogram(fill = "#67a9cf") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  xlab('Primary Education Type (private/public)') +
  ylab('Percentage')
```

## Health Expenditure

The variable `health.exp` has the information related to how much the surveyed
population is expending monthly in health, mainly expenses in health insurance, 
either social or private, which is mandatory for people in dependent or 
independent works.

```{r health_sum, echo=F}
t <- summary(subset(households, !is.na(health.exp))$health.exp)
pander(data.frame(Statistic = names(t), Value = as.vector(t)),
       caption = "Health Expenditure Summary Table")
```

The previous table states that the median expenditure is US$44, with a mean
of US$80.98, and a maximum amount of US$2279. How is the expenditure 
distributed? The following figure shows an histogram with this information.

```{r health, fig.align='center', echo=F}
ggplot(aes(health.exp), data = subset(households, !is.na(health.exp))) +
  geom_histogram(fill = "#67a9cf", binwidth = 10) +
  xlab('Health Expenditure (US$)') +
  ylab('Frequency')
```

---

As with other variables in previous sections, there are some outliers that 
prevent us to see the distribution in greater detail. So we create a new plot
with a logarithmic x axis scale.

```{r health_log, fig.align='center', echo=F}
ggplot(aes(health.exp), data = subset(households, !is.na(health.exp))) +
  geom_histogram(fill = "#67a9cf", binwidth = 0.2) +
  scale_x_continuous(trans = "log1p", breaks = c(5,50,100,200,1000,2000)) +
  xlab('Health Expenditure (US$)') +
  ylab('Frequency')
```

## Univariate Analysis and reflection

All of the variables explored in this section come from the `household` data 
set. Some of the variables are by household and other by individuals. The
most interesting features were the ones analysed: population age, educational
attainment, kinship, household tenure, household's income, educational 
institutions and health expenditure.

From the exploration, we found out that the Chilean population is young overall,
with 50% of the population under 32 years old. From the number of inhabitants
by household and the relationship of the dwellers with the houshold chief we
can also conclude that households are mainly composed by families with 
parents and childrend living together.

One of the most important conclusions is that the wealth distribution in Chile
seems to present high levels or inequality, with 50% of the households earning
less than US$1100 monthly, and the highest decil earning more than US$3610 and
concentranting around the 35% of the wealth.

# Bivariate/Multivariate Plot and Analysis.

## Population Pyramid

We can arrange a little bit more the plot built on 
[Household's inhabitants Age Distribution] and create a _population pyramid_:

```{r age_pyr, fig.align='center',fig.width=8, echo=F, warning=F}

ggplot() + 
  geom_histogram(
    aes(x = age, fill = gender),
    binwidth = 1, 
    data = subset(households, gender == "Women" & age >= 0)) + 
  geom_histogram(
    aes(x = age, fill = gender, y = ..count.. * -1.),
    binwidth = 1, 
    data = subset(households, gender == "Men" & age >= 0)) + 
  scale_fill_brewer(palette = "Paired") +
  scale_x_discrete(breaks = seq(0, 105, 5)) +
  scale_y_continuous(limits = c(-400, 400)) +
  xlab('Age') +
  ylab('Counts') +
  coord_flip()

gen_tab <- prop.table(table(households$gender))
```

The peaks seem to change for each gender! We can also notice that there are more
women (`r format(100 * gen_tab[[2]][1], digits = 3)`%) than men 
(`r format(100 * gen_tab[[1]][1], digits = 3)`%). Are the gender's average age
different?

```{r gender_age, fig.align='center', fig.width=8, echo=F}
ggplot(aes(age), data = households) + 
  geom_density(aes(color = gender), binwidth = 1) + 
  scale_color_brewer(palette = "Paired") +
  xlab("Age") +
  ylab("Density")
```

We see a difference in the average age for both gender, with males having an
overall younger population.

```{r gen_age_table, results='asis', echo=F}
age_gen <- subset(households, age >= 0, select = c("age", "gender"))
tc <- do.call(rbind, by(age_gen$age, age_gen$gender, summary))
pandoc.table(tc, split.cells = 5)
```

Let's test if the difference in statistical significant by using the Wilcoxon 
Rank Test:

```{r age_gen_test, results='asis', echo=F}
pander(wilcox.test(age ~ gender, age_gen), table.style = "multiline")
```

The test confirms that there the difference in age between genders is 
significant with p < 0.05.

## Educational Attainment and Age



We see a pick in the distribution at category 5, which correspond to the primary
education. This doesn't mean that most of the population only reach primary
school: we have not removed the school-age population. We could use two 
variables to subset the data and include only the population that is no longer
studying:

* `edu.finish`: indicates wheter the person did or not finish the educational
  attainment reported.
* `age`: we subset only the population over 30 years old, assuming that most
  of the people is no longer studying at that age.

```{r edu_att_fin, fig.align='center', echo=F, warning=F}

p1 <- ggplot(
  aes(edu.level, y = 100 * ..count.. / sum(..count..)),
  data = subset(households, studying == "No")) + 
  geom_histogram(fill = "#67a9cf") + 
  scale_x_discrete(labels = c(), drop = F) +
  scale_fill_brewer(palette = "Paired") +
  geom_vline(xintercept = 12.5) +
  coord_cartesian(ylim = c(0,32)) +
  ggtitle("Population not Studying (over 3 years old).")+
  xlab("Educational attainment") +
  ylab("Percentage")

p2 <- ggplot(
  aes(edu.level, y = 100 * ..count../sum(..count..)),
  data = subset(households, age >= 30 & studying == "No")) + 
  geom_histogram(fill = "#67a9cf") + 
  scale_x_discrete(labels = append(seq(0,15,1), "NA"), drop = F) +
  scale_fill_brewer(palette = "Paired") +
  geom_vline(xintercept = 12.5) +
  coord_cartesian(ylim = c(0,32))+
  ggtitle("Population not Studying (over 30 years old).") +
  xlab("Educational attainment") +
  ylab("Percentage")

grid.arrange(p1, p2, ncol = 1)
```


## Educational Attainment and Gender

We will study the distribution using an stacked histogram, so we can study at 
the same time if there are any significant differences between genders.

```{r eduatt_gen, fig.align='center', echo=F}

ggplot(aes(edu.level), data = households) + 
  geom_histogram(aes(fill = gender), position = 'dodge') + 
  scale_x_discrete(labels = append(seq(0,15,1), "NA")) +
  scale_fill_brewer(palette = "Paired")

edu_get <- 
  households %>% 
  group_by(gender, edu.level) %>% 
  summarize(count = n()) %>% 
  group_by(gender) %>% 
  mutate(gentot = sum(count))

ggplot(aes(edu.level), data = edu_get) + 
  geom_bar(
    aes(fill = gender, y = 100 * count / gentot), 
    position = 'dodge',
    stat='identity'
    ) +
  scale_x_discrete(labels = append(seq(0,15,1), "NA")) +
  scale_fill_brewer(palette = "Paired")
```

As shown in the [Introduction]



## Wage Income and Age

```{r age_inc, echo=F, fig.align='center', warning=F}
age_inc <- 
  ggplot(aes(age, dep.work.income), 
         data = subset(households, age > 0 & dep.work.income > 0)) +
  geom_jitter(alpha = 0.4) +
  scale_y_continuous(
    trans = "log1p", breaks = c(100, 500, 1000, 2500, 5000, 10000, 25000))

age_inc

sum_age_inc <-
  households %>%
  subset(age > 0 & dep.work.income > 0) %>%
  group_by(age) %>%
  summarise(mincome = median(dep.work.income))

age_inc + 
  geom_line(stat = 'summary', fun.y = median, color = "#67a9cf", size = 2) + 
  scale_x_continuous(limits = c(15, 80))
```

```{r inc_exp1, warning=F, echo=F, fig.align='center', message=F}
ggplot(aes(income.hh.av, expenses.tot),
       data = subset(households, household.head == 1)) +
  geom_point(alpha = 0.3) + 
  scale_x_log10(limits = c(40, 15000)) +
  scale_y_log10() +
  coord_cartesian(xlim = c(40,15000), ylim=c(100,10000)) +
  geom_smooth(method = 'lm')
```

## Wage Income, Age and Gender.

```{r gen_inc, fig.align='center', fig.width=8, echo=F}
households$age.group <- cut(households$age, seq(15, 81, 5), right = F)

ggplot(aes(age.group),
       data = subset(households, !is.na(dep.work.income) & !is.na(age.group))) +
  geom_boxplot(aes(y = dep.work.income, fill = gender)) + 
  scale_y_continuous(trans = 'log1p') + 
  scale_fill_brewer(palette = "Paired")

t1 <- dcast(subset(households, !is.na(dep.work.income)),
            age.group ~ gender,
            fun = median,
            value.var = 'dep.work.income')

w <- households %>%
  subset(!is.na(dep.work.income)) %>% 
  group_by(age.group, gender) %>% 
  summarise(inco = median(dep.work.income))

ggplot(aes(age.group), data = subset(w, !is.na(age.group))) + 
  geom_histogram(aes(y = inco, fill = gender), stat = 'identity',
                 alpha = 0.5, position = "identity", color = 'grey') + 
  scale_fill_brewer(palette = "Paired")
```

## Household's Income and Expenses

```{r inc_exp3, echo=F, fig.align='center', warning=F}
ggplot(aes(income.dec, income.hh.av - expenses.tot),
       data = subset(households, person.id == 1)) + 
  geom_hline(yintercept = 0) +
  geom_jitter(alpha = 0.1) +
  geom_boxplot(alpha = 0.5) +
  coord_cartesian(ylim = c(-1000, 5000)) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

ggplot(aes(income.dec, expenses.tot / income.hh.av),
       data = subset(
         households, person.id == 1 & expenses.tot / income.hh.av <= 10)
       ) + 
  geom_hline(yintercept = 1) +
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.6) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  scale_y_continuous(trans = 'log1p')

```


## Top Expenditures

What are households expending on?

```{r exp_cat1, fig.align='center', fig.width=10, warning=F, echo=F}

tabexp_sum <- 
  expenses %>% 
  group_by(description) %>% 
  summarise(expend = sum(expense))

topexp_by_sum <- as.character(
  as.data.frame(tabexp_sum[order(-tabexp_sum$expend)[2:33][-17][-14], ])[, 1])

topexp_sum_df <- subset(expenses, description %in% topexp_by_sum)

tabexp_count <- 
  expenses %>% 
  group_by(description) %>% 
  summarise(co = n())

topexp_by_count <- as.character(
  as.data.frame(tabexp_count[order(-tabexp_count$co)[1:31][-5], ])[, 1])

topexp_count_df <- subset(expenses, description %in% topexp_by_count)

tabexp_med <-
  topexp_sum_df %>% 
  group_by(description) %>% 
  summarise(med = median(expense))

topexp_by_med <- 
  as.character(as.data.frame(tabexp_med[order(-tabexp_med$med), ])[, 1])

ggplot(
  aes(factor(description, levels = topexp_by_med), expense),
  data = topexp_sum_df
  ) + 
  geom_jitter(alpha = 0.3, color = "#67a9cf") + 
  geom_boxplot(alpha = 0.8) + 
  scale_y_log10() + 
  scale_x_discrete(labels = 1:30) +
  xlab('Expenditure Category') +
  ylab('Expenditure Amount (US$)')
```

## Household Income Decil and Expenditures

```{r exp_cat2, fig.align='center', fig.width=10, warning=F, echo=F}
tabexp_med2 <- 
  topexp_count_df %>% 
  group_by(description) %>% 
  summarise(med = median(expense))

topexp_by_med_2 <- 
  as.character(as.data.frame(tabexp_med2[order(-tabexp_med2$med), ])[, 1])

ggplot(
  aes(factor(description, levels = topexp_by_med_2), expense), 
  data = topexp_count_df
  ) + 
  geom_jitter(alpha = 0.3, color = "#67a9cf") + 
  geom_boxplot(alpha = 0.8) + 
  scale_y_log10(limits = c(0.1, 10000)) + 
  scale_x_discrete(labels = 1:30) +
  xlab('Expenditure Category') +
  ylab('Expenditure Amount (US$)')
```



```{r fig.align='center', fig.width=9, echo=F}
t <- 
  left_join(
    expenses, 
    subset(households, person.id == 1,
           select = c(num.inhabitants, income.dec, home.id)), 
    by = 'home.id') %>% 
  subset(!is.na(income.dec))

expsum <- 
  t %>% 
  group_by(income.dec, d) %>% 
  summarise(total = sum(expense), median = median(expense), quant = n())

ggplot(aes(d, y = total), data = expsum) + 
  geom_bar(aes(fill = d), stat = 'identity') +
  scale_x_discrete(labels = '') +
  scale_fill_brewer(palette = "Paired") +
  facet_wrap( ~ income.dec)

ggplot(aes(income.dec, y = median), data = expsum) + 
  geom_bar(aes(fill = income.dec), stat = 'identity') +
  scale_x_discrete(labels = '') +
  scale_fill_brewer(palette = "Paired") +
  facet_wrap( ~ d)
```



D Code | Description
-------|------------------------------------------------------------------------
01     | Food and non-alcoholic beverages
02     | Alchoholic beverages, tobacco and narcotics
03     | Clothing and footwear
04     | Housing, water, electricity, gas and other fuels
05     | Furnishings, household equipment and routine household maintenance
06     | Health
07     | Transport
08     | Communication
09     | Recreation and culture
10     | Education
11     | Restaurants and hotels
12     | Miscellaneous goods and services

---

# Final Plots and Summary

## Plot One

```{r cum_inc, echo=F, fig.align='center'}
houseinc$per <- houseinc$income.hh.av.rent / sum(houseinc$income.hh.av.rent)

cuminc <- ecdf(cumsum(houseinc[order(houseinc$per),'per']))

ggplot(houseinc[order(houseinc$per),], aes(cumsum(per) * 100.)) + 
  stat_ecdf(show_guide = FALSE) + 
  geom_hline(
    aes(yintercept = cuminc(0.25),
        color = format(100 * cuminc(0.25), digits = 3)),
    show_guide = TRUE) +
  geom_hline(
    aes(yintercept = cuminc(0.5), 
        color = format(100 * cuminc(0.5), digits = 3)), 
    show_guide = TRUE) +
  geom_hline(
    aes(yintercept = cuminc(0.75), 
        color = format(100 * cuminc(0.75), digits = 3)), 
    show_guide = TRUE) +
  coord_cartesian(xlim = c(0, 100)) + 
  scale_color_manual(name = "Pop. %", values = c("red", "#67a9cf", "blue")) +
  scale_y_continuous(labels = percent)
```

So, while the mean household income is 
US$`r summary(houseinc$income.hh.av.rent)[4]` the median is at 
US$`r summary(houseinc$income.hh.av.rent)[3]`. 

## Plot Two

```{r eduatgen2, fig.align='center', fig.width=10, fig.height=7, echo=F}

households$age.group <- cut(
  households$age, seq(1, 101, 10),
  labels = c("0-9 years", "10-19 years", "20-29 years", "30-39 years",
             "40-49 years", "50-59 years", "60-69 years", "70-79 years",
             "80-89 years", "90-99 years"),
  right = F)

households$edu.cat <- ifelse(
  as.numeric(households$edu.level) > 11, "Tertiary", "No.Tertiary")

ctable <- 
  as.data.frame(
  with(households, table(age.group, edu.cat, gender))
  ) %>% 
  group_by(age.group, gender) %>% 
  mutate(Total = sum(Freq))

ggplot(aes(edu.cat, y = 100 * Freq / Total), data = ctable) + 
  geom_bar(aes(fill = gender), stat = "identity", position = "dodge") + 
  scale_x_discrete(labels = c('No Tertiary', 'Tertiary')) + 
  xlab("Educational Attainment") +
  ylab("Percentage by age group") +
  scale_fill_brewer(palette = "Paired") +
  facet_wrap( ~ age.group)
```

## Plot Three

```{r fig.align='center', fig.width=9, echo=F}
expsum <- 
  expsum %>% 
  group_by(income.dec) %>% 
  mutate(suma = sum(total))

ggplot(aes(income.dec, y = 100 * total / suma), data = expsum) + 
  geom_bar(aes(fill = income.dec), stat = 'identity') + 
  scale_x_discrete(labels = '') +
  scale_fill_brewer(palette = "Paired") +
  facet_wrap( ~ d)
```

# Reflection

